Adaptive Sequential Model Selection

### 2015-06-30 paper.tex
rob changes

\begin{abstract}
  Many model selection algorithms produce a ``path'' of fits specifying a sequence of increasingly complex models. Given such a sequence and the data used to produce them, we consider the problem of choosing the least complex model that is not falsified by the data. Extending the selected-model tests of \citet{fithian2014optimal}, we construct $p$-values for each step in the path, accounting for the fact that the model path is determined adaptively using the data. In the case of linear regression, our $p$-values improve on the power of the spacings test of \citet{taylor2014exact}, often dramatically.

@@

Many model selection procedures produce a path of fits

... generates a sequence of increasingly complex models. Our goal is to choose the simplest model that is not falsified by the available data.

... and find the index of the smallest adequate model --- that is, the smallest model that cannot be falsified given available evidence.

\begin{example}[Forward-Stepwise Linear Regression with the LASSO]
  \citet{taylor2014exact}
\end{example}

\begin{example}[The LARS Algorithm in Regression]
  \citet{taylor2014exact}
\end{example}

\begin{example}[Ever-Active Path in $\ell_1$-Regularized Methods]
  \citet{taylor2014exact}
\end{example}

\begin{example}[Principal Components Analysis]
  As a second motivating example, consider model selection for   principal components analysis. In that case we are given a data matrix $X \in \R^{n\times d}$, with which we form a sample covariance matrix
\[
S = \frac{1}{n-1} \sum_{i=1}^n(x_i - \bar x)^2
\]
The first $d$ principal component loadings are the first $d$ eigenvectors of $S$, which call $u_1,\ldots, u_d$. These induce a sequence of nested Wishart models:
\[
M_0 \sub M_1 \sub \cdots \sub M_d
\]
in which
\begin{equation}
  M_k:\; (n-1) S \sim W_d\left(\lambda_0 I_d + \sum_{i=1}^k     \lambda_i u_i u_i', \;\;\; n-1\right).
\end{equation}
This problem was studied in \citet{choi2014selecting}.
\end{example}

@@

\subsection{Which Null Hypothesis Should We Test?}

@@

As usual, we have reduced computation of the $p$-value at step $k$ to a sampling problem. Whatever our test statistic at step $k$, we must compute its distribution under the null model $M_{k-1}$.

Happily, the polytope to condition on has only $2p$ constraints in forward stepwise regression, as we see next.

@@

@@

Next we compare several model selection procedures in simulation. We simlate from a linear regression model with $n=100$ observations and $p=40$ variables. The design matrix $X\in\R^{n\times p}$ is a random Gaussian design with pairwise correlations of 0.3 between predictor variables.

@@

\begin{figure}
  \centering
  % source code: ??? for simulation, sparseSim.R for plotting
  \includegraphics[width=.8\textwidth]{figs/simulation_snr_5_alpha_05_null_false.pdf}
  \caption{CDFs of saturated-model (black), selected-model (red), and nominal (blue) $p$-values in the simulation of Section~\ref{sec:sparseReg}, conditional on testing a false null hypothesis at step $k$. The selected-model test is much more powerful than the saturated-model test. The nominal test appears powerful, but is not a correct $p$-value.}
  \label{fig:simulation_null_false}
\end{figure}

\begin{figure}
  \centering
  % source code: ??? for simulation, sparseSim.R for plotting
  \includegraphics[width=.8\textwidth]{figs/simulation_snr_5_alpha_05_null_true.pdf}
  \caption{CDFs of saturated-model (black), selected-model (red), and nominal (blue) $p$-values in the simulation of Section~\ref{sec:sparseReg}, conditional on testing a true null hypothesis at step $k$. The selected-model test is much more powerful than the saturated-model test. The nominal test is badly anti-conservative.}
  \label{fig:simulation_null_true}
\end{figure}

\begin{figure}
  \centering
  % source code: ??? for simulation, sparseSim.R for plotting
  \includegraphics[width=.8\textwidth]{figs/simulation_snr_5_alpha_05_noise_var.pdf}
  \caption{CDFs of saturated-model (black), selected-model (red), and nominal (blue) $p$-values in the simulation of Section~\ref{sec:sparseReg}, conditional on the event that the variable added at step $k$ is a noise variable in the full model. Here, none of the methods produce uniform $p$-values. The null hypothesis is false in most cases and so --- in our model-centric point of view --- rejection is the desired outcome.}
  \label{fig:simulation_noise_var}
\end{figure}

>>>>

\begin{abstract}
  Many model selection algorithms produce a ``path'' of fits specifying a sequence of increasingly complex models. Given such a sequence and the data used to produce them, we consider the problem of choosing the least complex model that is not falsified by the data. Extending the selected-model tests of \citet{fithian2014optimal}, we construct $p$-values for each step in the path, accounting for the fact that the model path is determined adaptively using the data. In the case of linear regression, our $p$-values improve on the power of the spacings test of \citet{taylor2014exact}, often dramatically.
  We also show that the selected models test  bears a close relationship to parametric bootstrap sampling,

@@

Many model selection procedures produce  a sequence of increasingly complex models. Often our goal is to choose the simplest model that is not falsified by the available data.
In other words, we seek  the index of the smallest adequate model ---  the smallest model that cannot be falsified given available evidence.
Examples of such model selection procedures include  forward stepwise linear regression, least angle regression (LAR) and  the ever-active path in $\ell_!$-regularized models .
All of these were studied in \citet{taylor2014exact}. Another, different, example is principal components analysis in which we try to determine the correct rank of the
covariance matrix. This problem was studied in \citet{choi2014selecting}.

A natural way carry out this model selection is as follows: we compute a set  of sequential p-values   at each step of the path of models and then feed them into a stopping rule
that provides guaranteed control of the false discovery rate (FDR) or a similar quantity. Recently  \citet{gsell2013sequential} and \citet{li2015accumulation} proposed sequential stopping rules of this kind.
Both sets of rules require that once we have reached the correct model in our path, the p-values in subsequent steps are uniform (or sub-uniform)
and independent.
In this paper we develop a theoretical framework for the construction of sequential p-values  that are both (sub)-uniform and independent under the correct model,
and show how to construct them  in forward stepwise regression, LAR and the ever-active path in $\ell_1$-regularized models.
In the former case, the test is simply a sequential selective version the max-$t$ test of ~\citet{bujaxxxxmaxT}, based one the theory developed \citet{fithian2014optimal}.

Table \ref{tab:diab} shows an example: using the diabetes data taken from \cite{LARS}.
These data consist of  observations on $442$ patients, with the response of interest being a quantitative measure of disease progression one year after baseline.
There are ten baseline variables---age, sex, body-mass index, average blood pressure,
and six blood serum measurements---plus quadratic terms, giving a total of 64 features.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline\\
 Step& Predictor& Naive pvalue& Saturated pvalue& Max-t pvalue\cr
  \hline\\
1 & 3 & 0.000 & 0.000 & 0.000 \\ 
  2 & 9& 0.000 & 0.000 & 0.000 \\ 
  3 & 4 & 0.000 & \color{red}{0.054} & 0.000 \\ 
  4 & 20& 0.000 & { 0.333} & 0.003 \\ 
  5 & 37& 0.002 & 0.756 & 0.043 \\ 
  6 & 7 & 0.003 & 0.254 & 0.037 \\ 
  7 & 2& 0.000 & 0.003 & 0.000 \\ 
  8 & 19 & 0.020 & 0.028 &  \color{red}{0.343} \\ 
  9 & 11 & 0.112 & 0.548 & 0.937 \\ 
  10 & 52& 0.214 & 0.911 & 0.907 \\ 
  11 & 57& 0.049 & 0.000 & 0.954 \\ 
  12 & 24& 0.222 & 0.000 & 0.996 \\ 
  13 & 5& 0.183 & 0.054 & 0.789 \\ 
  14 & 49 & 0.158 & 0.333 & 1.000 \\ 
  15 & 30 & 0.212 & 0.756 & 1.000 \\ 
  16 & 46 & \color{red}{0.159} & 0.254 & 0.000 \\ 
  17 & 59 & 0.270 & 0.003 & 0.004 \\ 
  18 & 33& 0.116 & 0.028 & 1.000 \\ 
  19 & 62 & 0.081 & 0.548 & 0.668 \\ 
  20 & 27 & 0.170 & 0.911 & 1.000 \\

   \hline\\
\end{tabular}
\caption[tab:diab]{\em Forward stepwise regression for the diabetes data: naive p-values from forward stepwise regression, 
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper. The red annotation indicates the stopping step
prescribed by {\tt forwardStop}. For the max-t p-values, this gives exact FDR control at the 10\% level.}
\label{tab:diab}
\end{table}
The table shows the naive p-values from forward stepwise regression, that do not account for the selection, as well as the
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper.
The latter two sets of p-values test someone different hypotheses:  the p-values from the saturated model test whether the partial regression of the predictor just added is zero,
what we call the ``incremental null''
In contrast, the max-t pvalues test whether the current model is adequate, that is, all predictors not in the model  have coefficients equal to zero,
the ``complete null''. (The issue of what hypothesis to test is discussed in section \ref{sec:whichnull}.)
Now it turns out that the max-t test if often more powerful that the saturated model test, and importantly, the max-t pvalues are exactly
independent after we reach the point the model sequence when the complete null is true. This independence means that we can apply
sequential FDR exact stopping rules, such as ``forwardStop'' (\citet{sell2013sequential}). When applied to the max-t pvalues at FDR 0.10, {\tt forwardStop}
suggested stopping at step 8. In constant, the saturated lvalues gave stopping a point of 3, with no guarantee of FDR control
due to the lack of independence of the p-values.

An outline of this paper is as follows:

OUTLINE

%\begin{example}[Forward-Stepwise Linear Regression]
 % \citet{taylor2014exact}
%end{example}

%\begin{example}[The LARS Algorithm in Regression]
%  \citet{taylor2014exact}
%\end{example}

%\begin{example}[Ever-Active Path in $\ell_1$-Regularized Methods]
 % \citet{taylor2014exact}
%\end{example}

%\begin{example}[Principal Components Analysis]
%  As a second motivating example, consider model selection for   principal components analysis. In that case we are given a data matrix $X \in \R^{n\times d}$, with which we form a sample %covariance matrix
%\[
%S = \frac{1}{n-1} \sum_{i=1}^n(x_i - \bar x)^2
%\]
%The first $d$ principal component loadings are the first $d$ eigenvectors of $S$, which call $u_1,\ldots, u_d$. These induce a sequence of nested Wishart models:
%\[
%M_0 \sub M_1 \sub \cdots \sub M_d
%\]
%n which
%\begin{equation}
 % M_k:\; (n-1) S \sim W_d\left(\lambda_0 I_d + \sum_{i=1}^k     \lambda_i u_i u_i', \;\;\; n-1\right).
%\end{equation}
%This problem was studied in \citet{choi2014selecting}.
%\end{example}

@@

\subsection{Which Null Hypothesis Should We Test?}
\label{sec:whichnull}

@@

As usual, we have reduced computation of the $p$-value at step $k$ to a sampling problem. Whatever our test statistic at step $k$, we must compute its distribution under the null model $M_{k-1}$. We give details of our  strategy here for the max-t test in forward stepwise regression.

Given a model $M_{k-1}$   we need to sample from 
\begin{eqnarray}
L(X'_jY | E_{k-1},X'_{E_{k-1}})
\label{eqn:maxtsamp}
\end{eqnarray}
with $Y\sim N(\mu, I \cdot \sigma^2)$.
The current model fit is $\proj_{E_{k-1}} Y$ and is is easy to see that (\ref{eqn:maxtsamp})  is equivalent to sampling
\begin{eqnarray}
Y^*=\proj_{E_{k-1}}) Y + (I- \cdot \proj_{E_{k-1}}) Z; \;Z\sim N(0,I\sigma^2)
\label{eq:boot}
\end{eqnarray}
and then conditioning on the event $E_{k-1}(Y^*)= E_k(Y)$ (a polyhedron).
In other words, we need to condition on data vectors  $Y*$ that yields the same
sequence of variables and coefficient signs up to step $k-1$. This is the challenging aspect of this computation.

Note that  (\ref{eq:boot}  has the form of  (conditional) parametric bootstrap sampling, in which the bootstrap residuals projected into the subspace
orthogonal to the current fit. In similar fashion, one might use a instead a nonparametric version of (\re{eq:boot}), sampling
with replacement from the model residuals and then projecting them orthogonal to the current model.

Figure \ref{fig:comparison} shows and example with $n=100,p=20$ and one strong signal, computing the max-t pvalue at Step 3.
The ``exact'' answer  was determined from bootstrap sampling with $20,000$ replications.
The Figures show the bias, variance and MSE of the estimates from accept/reject and hit and run, using between 1000 and 8000 samples, averaged
over 10 different starting seeds.
\begin{figure}[htp]
\centering
  \includegraphics[width=\textwidth]{figs/comparison.pdf}
  \caption{\em Simulated example with $n= 100, p=20$ and one strong signal, computing the max-t pvalue at Step 3.  Shown are bias, variance and MSE of the estimates from accept/reject and hit and run, using between 1000 and 8000 samples, averaged
over 10 different starting seeds.
}
  \label{fig:comparison}
\end{figure}
We see that accept/reject is far more efficient, yielding a smaller MSE with 1000 samples then hit and run gives for  8000 samples.
Hence we use accept/reject for as many steps of the sequence is possible,  until the point where the acceptance rate is too low to make is practical.

In detail, we employ a hybrid computational strategy; we use accept-reject sampling, and do so as long as we obtain at least 300 (say) samples in the first set
of $B$ realizations ($75,000$ in our code)  This is an effective approach until we have reached a point in the model path where all predictors have little or no signal.
When the acceptance rate gets too low, we switch to a hit-and run Monte Carlo approach. We exploit the fact that each move along a random direction 
has a truncated Gaussian distribution whose truncation limits are easily computed from the properties of the polyhedral. Unlike accept-reject, sampling
which produces independent samples,  the hit and run method produces dependent samples. Hence we must run it for longer with an initial burn in period,
and hope that it has mixed sufficiently well.

The task is made much simpler by the fact that the selection event can we written as $S_M=\{y: \Gamma y \geq u\}$.
Hence we don't need to run the entire forward stepwise procedure on $y^*$: instead, we pre-compute $\Gamma$ and $u$ and then check
if $\Gamma y^* \geq u$. Further, since $\Gamma$ has on the  order of $2pk$ rows after $k$ steps,  we can save 
substantially in computation by checking the row-wise inequalities starting from the last row, where the inequality  it is most likely to fail.

\subsection{Reducing the number of constraints}
Happily, it turns out that the conditioning polytope can be expressed with only $2p$ constraints in forward stepwise regression, as we show here.
This increases the speed of both accept/reject and hit andRun sampling for our problem.

@@

\subsection{An analytic approximation for the max-t test}
 Suppose we have taken $k$ steps of forward stepwise regression with $p$ predictors and let the selected model  
 be $E$. Suppose we choose predictor $j^*$ at the next step.

Let $T_j=\eta_j^Ty, j \notin M$ be the partial regression coefficients for all predictors not in $E$,
and $(v_j^-, v_j^+)$ be the limits derived from the polyhedral lemma applied to $(E, \eta_j)$  (See e.g. \citet{spacings}).
These are the upper and lower bounds for the functional $\eta_j^Ty$ implied by the event that model $E$ is selected from the data $y$.
We chose the sign of $\eta_j$ so that $\eta_j^Ty$ is  $\geq 0$.

 Let $F_{\mu,\sigma^2}^{a,b}(t)$ be the  cumulative distribution function of the truncated normal.
 Finally let $T=|T_{j^*}|$.
Then by assuming independence, a  simple approximation to the max-t value is given by
\begin{equation}
1-\prod_{j\notin E} \{F_{0,||\eta||^2}^{v_j^-, v_j^+}(T)-   F_{0,||\eta||^2}^{v_j^-, v_j^+}(-T) \}
\label{eqn:approxpv}
\end{equation}
Figure \ref{fig:approxpv} shows an example with $n=50$ and $p=10$ features having pairwise correlation 0.5.
There are two moderate signals.
Shown are the p-values from the ``exact'' numerical computation (with $100,000$ bootstrap samples)
and the approximation based on (\ref{eqn:approxpv}).
\begin{figure}[htp]
\centering
  \includegraphics[width=3in]{figs/approxpv.pdf}
  \caption{Exact max-t pvalues and approximation based on (\ref{eqn:approxpv}), for a simulated example
  with $n=50, p=10$.}
  \label{fig:approxpv}
\end{figure}
The approximation seems quite good. By Sidak's lemma, we might conjecture that the approximation is conservative, but we have no proof of this.

@@

Next we compare several model selection procedures in simulation. We simulate from a linear regression model with $n=100$ observations and $p=40$ variables. The design matrix $X\in\R^{n\times p}$ is a random Gaussian design with pairwise correlations of 0.3 between predictor variables.

@@

\begin{figure}
  \centering
  % source code: ??? for simulation, sparseSim.R for plotting
  \includegraphics[width=.8\textwidth]{figs/simulation_snr_5_alpha_05_null_false.pdf}
  \caption{\em CDFs of saturated-model (black), selected-model (red), and nominal (blue) $p$-values in the simulation of Section~\ref{sec:sparseReg}, conditional on testing a false null hypothesis at step $k$. The selected-model test is much more powerful than the saturated-model test. The nominal test appears powerful, but is not a correct $p$-value.}
  \label{fig:simulation_null_false}
\end{figure}

\begin{figure}
  \centering
  % source code: ??? for simulation, sparseSim.R for plotting
  \includegraphics[width=.8\textwidth]{figs/simulation_snr_5_alpha_05_null_true.pdf}
  \caption{\em CDFs of saturated-model (black), selected-model (red), and nominal (blue) $p$-values in the simulation of Section~\ref{sec:sparseReg}, conditional on testing a true null hypothesis at step $k$. The selected-model test is much more powerful than the saturated-model test. The nominal test is badly anti-conservative.}
  \label{fig:simulation_null_true}
\end{figure}

\begin{figure}
  \centering
  % source code: ??? for simulation, sparseSim.R for plotting
  \includegraphics[width=.8\textwidth]{figs/simulation_snr_5_alpha_05_noise_var.pdf}
  \caption{\em CDFs of saturated-model (black), selected-model (red), and nominal (blue) $p$-values in the simulation of Section~\ref{sec:sparseReg}, conditional on the event that the variable added at step $k$ is a noise variable in the full model. Here, none of the methods produce uniform $p$-values. The null hypothesis is false in most cases and so --- in our model-centric point of view --- rejection is the desired outcome.}
  \label{fig:simulation_noise_var}
\end{figure}
